### 2021-01-23 paper/intro.tex
Fleshing out Sec 1.1

This has led to \emph{relaxed memory models}, in which the requirement of
sequential consistency is weakened to only apply \emph{per-location} and not globally
over the whole progfram. This allows executions which
are inconsistent with program order, such as:
  \begin{gather*}
    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
    \PAR
    \THREAD{r\GETS y\SEMI s\GETS x}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=of wy2}
        \event{rx0}{\DR{x}{0}}{right=of ry2}
        \rf[out=20,in=160]{wy2}{ry2}
        \rf[out=15,in=165]{wx0}{rx0}
        \wk{wx0}{wx1}
        \wk[out=200,in=340]{rx0}{wx1}
      \end{tikzinline}}
  \end{gather*}
In such models, the causal order between events is important,
and includes control and data dependencies, to avoid
paradoxical ``out of thin air'' examples such as:
  \begin{gather*}
    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \FI}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{rx1}{wy1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
This candidate execution forms a cycle in causal order, so is disallowed,
but this depends crucially on the control dependency
from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
is acyclic and hence allowed. For example dropping the control dependency
results in:
  \begin{gather*}
    \THREAD{r\GETS x\SEMI y\GETS1}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
Unfortunately, while a simple syntactic approach to dependency calculation
suffices for hardware models, it is not preserved by common compiler
optimizarions. For example, if we calculate control dependencies syntactically,
then there is a dependency from $(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate execution:
  \begin{gather*}
    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \ELSE y\GETS1 \FI}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{rx1}{wy1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
An optimizing compiler might lift the assignment $y\GETS1$ out of the conditional,
thus removing the control dependency.

@@

\end{itemize}
In this paper, we will focus on logical models, as those are compositional,
and align well with existing models of sequential composition.

\subsection{Sequential composition}

>>>>

This has led to \emph{relaxed memory models}, in which the requirement of
sequential consistency is weakened to only apply \emph{per-location} and not globally
over the whole progfram. This allows executions which
are inconsistent with program order, such as:
  \begin{gather*}
    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
    \PAR
    \THREAD{r\GETS y\SEMI s\GETS x}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=of wy2}
        \event{rx0}{\DR{x}{0}}{right=of ry2}
        \rf[out=20,in=160]{wy2}{ry2}
        \rf[out=15,in=165]{wx0}{rx0}
        \wk{wx0}{wx1}
        \wk[out=200,in=340]{rx0}{wx1}
      \end{tikzinline}}
  \end{gather*}
In such models, the causal order between events is important,
and includes control and data dependencies, to avoid
paradoxical ``out of thin air'' examples such as:
  \begin{gather*}
    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \FI}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{rx1}{wy1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
This candidate execution forms a cycle in causal order, so is disallowed,
but this depends crucially on the control dependency
from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
is acyclic and hence allowed. For example dropping the control dependency
results in:
  \begin{gather*}
    \THREAD{r\GETS x\SEMI y\GETS1}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
Unfortunately, while a simple syntactic approach to dependency calculation
suffices for hardware models, it is not preserved by common compiler
optimizarions. For example, if we calculate control dependencies syntactically,
then there is a dependency from $(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate execution:
`  \begin{gather*}
    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \ELSE y\GETS1 \FI}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{rx1}{wy1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
An optimizing compiler might lift the assignment $y\GETS1$ out of the conditional,
thus removing the control dependency.

@@

\end{itemize}
In this paper, we will focus on logical models, as those are compositional,
and align well with existing models of sequential composition.
The heart of the model of~\cite{???} is to add logical preconditions
to events, which are introduced by store (modelling data dependencies)
or conditionals (modelling control dependencies):
  \begin{gather*}
    \IF{q \le r} \THEN z\GETS s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wz1}{(q {\le} r) \land (s{=}1) \mid \DW{z}{1}}{}
      \end{tikzinline}}
  \end{gather*}
and are discharged by being ordered with a read event:
  \begin{gather*}
    r\GETS x\SEMI s\GETS y\SEMI \IF{r\le r} \THEN z\GETS s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{ry1}{\DR{y}{1}}{right=of rx1 }
        \event{wz1}{(s{=}1) \limplies (r {\le} r) \land (s{=}1) \mid \DW{z}{1}}{right=of ry1}
        \po{ry1}{wz1}
      \end{tikzinline}}
  \end{gather*}
Note that there is dependency order from $(\DR{y}{1})$ to $(\DW{z}{1})$
so the precondition for $(\DW{z}{1})$ only has to be satisfied assuming the hypothesis
$(s=1)$. There is no matching order from $(\DR{x}{1})$ to $(\DW{z}{1})$
which is why we do not assume the hypothesis $(r=1)$. The precondition on
$(\DW{z}{1})$ is a tautology, and so can be elided:
  \begin{gather*}
    r\GETS x\SEMI s\GETS y\SEMI \IF{r\le r} \THEN z\GETS s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{ry1}{\DR{y}{1}}{right=of rx1 }
        \event{wz1}{\DW{z}{1}}{right=of ry1}
        \po{ry1}{wz1}
      \end{tikzinline}}
  \end{gather*}
\subsection{Sequential composition}
